Two Dimensional CSDA at the scale of neuronal tissue.

### 2015-07-03 Reportes/CSDAattheSomaticLayersScale.tex
corrigiendo textos

\begin{abstract}
As high density electrode arrays become cheaper and more common, 
electrophysiological measurements of activity in neural tissue approach
the scale of the underlying structures. The electrodes measure
directly the Local Field Potential, but it is aknowledged that the
Current Source Density has more resolving power and distinguishes
clearly between the loci of outgoing current and ingoing currents.
 The role of the conductivity tensor is rarely taken into account
in the derivation of the CSD, and this could potentialy lead to
spurious distributions of the sources and sinks- 
In dense neural tissue, such as the
stratum piramidale layer of the hippocampus, clear
deviations from average conductivity have been measured. 
Here we show a possible qualitative treatment of the conductivity
across the layers of the hippocampus, and compare the results
when this inhomogenity is taken into account and when its not.
\end{abstract}

@@

As Bédard and Destexhe pointed out in \cite{Bedard11}, the CSDA, as it
is performed usually, rests on
assumptions which could leave important features out. One of these 
assumptions is that the extracellular medium has very homogeneous 
electrical propierties. It has been shown not to be
the case for dense neuronal layers such as the piramidal
strata in the hippocampus \cite{Holsheimer87, Lopez01, TrevinoPersonal}.
The evidence mostly points out that resistivity grows 
considerably around thick somatic layers.
The following linear relationship is a good aproximation for
most media at convencional regimes: 
\begin{equation}
\Jd=\cond \EF,
\end{equation}
that is, the current density $\Jd$ is the result of the
local electric field $\EF$ and the local conductivity, $\cond$.
We usually take this quantity as constant, 
at least over macroscopic scales.
The conductivity in inhomogeneous media can be rarely
simplified this much, as it can depend both from the direction
of the fields and currents, and vary from point to point. 
Then the above expression is just shorthand notation for the
explicit  \emph{tensorial} relationship:
\begin{equation}
  \begin{pmatrix}
    J_x \\
    J_y \\
    J_z
  \end{pmatrix}
  =
   \begin{pmatrix}
     \sigma_{xx}E_x+ \sigma_{xy}E_y+\sigma_{xz}E_z \\
     \sigma_{xy}E_x+ \sigma_{yy}E_y+\sigma_{yz}E_z \\
     \sigma_{xz}E_x+ \sigma_{yz}E_y+\sigma_{zz}E_z \\
  \end{pmatrix},
\end{equation}
where we allready used the fact that $\cond$ is simmetric when
expressed on Cartessian coordinates.
We assume that the electric field can be obtained as the
gradient of the LFP, $\phi$:
\begin{equation}
  -\nabla \phi=\EF
\end{equation}
This LFP is the quantity that the microelectrode arrays (MEAs) permits us
to measure on the extracelular medium.
To obtain the sources and sink densities, we use Gauss Divergence Theorem,
\begin{equation}
\nabla \cdot \Jd =I_m
\end{equation}
The usual definition of $-\partial_t \rho =I_m$ has been made here. 
We perform the divergence operation on the current density, \emph{not}
on the electric field. When the medium is isotropic and homogeneous
both derivations coincide within a positive factor.
This makes sense when we are dealing with macroscopic quantities,
that is, that the are at last one order of magnitude larger
than the underlying physical structures which could cause 
inhomogenities, but as our measurements get finner and finner,
this treatment begins to be faulty. 
In the next section we investigate the role of the inhomogenities 
of the conductivity on current density
($\Jd$) and current source density ($I_m$) at scales around
the $20\mu m$.

>>>>

\begin{abstract}
As high density electrode arrays become cheaper and more common, 
electrophysiological measurements of activity in neural tissue approach
the scale of the underlying structures. The electrodes measure
directly the Local Field Potential, but it is aknowledged that the
Current Source Density has more resolving power and distinguishes
clearly between the loci of outgoing current and ingoing currents.
 The role of the conductivity tensor is rarely taken into account
in the derivation of the CSD, and this could potentialy lead to
spurious distributions of the sources and sinks- 
In dense neural tissue, such as the
stratum piramidale layer of the hippocampus, clear
deviations from average conductivity have been measured. 
Here we show a possible qualitative treatment of the conductivity
across the layers of the hippocampus based on coarse measurements
and general smoothness arguments.
\end{abstract}

@@

As Bédard and Destexhe pointed out in \cite{Bedard11}, the CSDA, as it
is performed usually, rests on
assumptions which could leave important features out. One of these 
is that the extracellular medium has very homogeneous 
electrical propierties. This may depend on the scale in which measurements
take place and the size of the neuronal structures involved. We shall
focus our atention on the hippocampus of the rat and mouse
 and its typical sizes in what follows.

The CSDA is the application of charge conservation theorem
for ionic dilutions in the extracelular medium for neuronal
tissue. CSD are obtained either directly through application
of a numeric Laplacian operator directly to LFP data \cite{Nicholson75}, 
or through  more refinated
techniques, called inverse CSDA \cite{Pettersen2006, Leski2011} or kernel CSDA 
\cite{Potworowsky2011}. There the conductivity $\cond$ is taken as 
 a numerical constant, that is:
\begin{equation}
-\nabla^2 \Phi = \frac{\Jd}{\cond}. 
\end{equation}
Here $\Phi$ is the LFP and $\Jd$ the current density. 
This is indeed a reasonable treatment at scales larger than $100\mu m$,
where each electrode measures averaged contributions from many different
densities of activity and structures. Even more, data are 
generaly smoothed out for presentation, and this contributes to the
averaging process.

On the other hand, it has been shown that at 
dense neuronal layers such as the piramidal
strata in the hippocampus the conductivity (or equivalently,
the resistivity) varies greatly around its mean value along
an axis which transverses said structure
\cite{Holsheimer87, Lopez01, TrevinoPersonal}. 
The evidence points out that resistivity grows 
considerably around the thick somatic layers. These layers 
have thickness between $40\mu m$ to $100 \mu m$.
The following linear relationship is a good aproximation for
most media at convencional regimes: 
\begin{equation}
\Jd=\cond \EF,
\end{equation}
that is, the current density $\Jd$ is the result of the
local electric field $\EF$ and the local conductivity, $\cond$.
We usually take this quantity as constant, 
at least over macroscopic scales.
The conductivity in inhomogeneous media can be rarely
simplified this much, as it can depend both from the direction
of the fields and currents, and vary from point to point. 
Then the above expression is just shorthand notation for the
explicit  \emph{tensorial} relationship:
\begin{equation}
  \begin{pmatrix}
    J_x \\
    J_y \\
    J_z
  \end{pmatrix}
  =
   \begin{pmatrix}
     \sigma_{xx}E_x+ \sigma_{xy}E_y+\sigma_{xz}E_z \\
     \sigma_{xy}E_x+ \sigma_{yy}E_y+\sigma_{yz}E_z \\
     \sigma_{xz}E_x+ \sigma_{yz}E_y+\sigma_{zz}E_z \\
  \end{pmatrix},
\end{equation}
where we allready used the fact that $\cond$ is simmetric when
expressed on Cartessian coordinates.
We assume that the electric field can be obtained as the
gradient of the LFP, $\phi$:
\begin{equation}
  -\nabla \phi=\EF
\end{equation}
This LFP is the quantity that the microelectrode arrays (MEAs) permits us
to measure on the extracelular medium.
To obtain the sources and sink densities, we use Gauss Divergence Theorem,
\begin{equation}
\nabla \cdot \Jd =I_m
\end{equation}
The usual definition of $-\partial_t \rho =I_m$ has been made here. 
We perform the divergence operation on the current density, \emph{not}
on the electric field. When the medium is isotropic and homogeneous
both derivations coincide within a positive factor.
This makes sense when we are dealing with macroscopic quantities,
that is, that the are at last one order of magnitude larger
than the underlying physical structures which could cause 
inhomogenities, but as our measurements get finner and finner,
this treatment begins to be faulty. 
In the next section we investigate the role of the inhomogenities 
of the conductivity on current density
($\Jd$) and current source density ($I_m$) at scales around
the $20\mu m$.
